---
title: "`dplyr` workshops"
author: "Natalia Potocka"
date: "March 16, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = 'hide',message = FALSE)
```


```{r}
library(dplyr)
library(nycflights13)

data(flights)

glimpse(flights)
```

`dplyr::%>%` passes object on left hand side as first argument (or `.` argument) of function on righthand side. "Piping" with `%>%` makes code more readable, 

```r
x %>% f(y)
```
is the same as 
```r
f(x, y)
y %>% f(x, ., z)
```
is the same as 
```r
f(x, y, z)
```

e.g.
```{r}
sum(c(1,2,3))
c(1,2,3) %>% sum()
```

A shortcut in RStudio for `%>%` is **CTRL + SHIFT + M** or **CMD + SHIFT + M**.


## `dplyr`'s main functions:

* `select` - select columns/variables
* `arrange` - sort the observations
* `mutate` - create a new column/variable
* `filter` - choose rows that satisfy certain condition
* `summarise` - summarise
* `group_by` - count in goups
* `join` - join two data sets

[Cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) or Help -> Cheatsheets -> Data manipulation with dplyr, tidyr


## `select`

Selecting one column:

```{r}
flights %>% 
  select(dep_delay)
```

Selecting more than one column:

```{r}
flights %>% 
  select(dep_delay, dep_time)
```

Selecting columns that contains `dep` in the name


```{r}
flights %>% 
  select(contains("dep"))
```


**Exercise 1.** Choose columns that have `arr` int the name.

## `arrange`

Sorting columns by departure delay

```{r}
flights %>% 
  arrange(dep_delay)
```

The same but choosing only two columns

```{r}
flights %>% 
  select(dep_delay, arr_delay) %>% 
  arrange(dep_delay)
```

The same but sorting desceding

```{r}
flights %>% 
  select(dep_delay, arr_delay) %>% 
  arrange(desc(dep_delay))
```

**Exercise 2.** Choose two columns: `distance` and `arr_delay` and sort data by distance.

## `mutate`

Create new variable indicating whether there was a delay or not

```{r}
flights %>% 
  mutate(if_arr_delay = (arr_delay > 0)) %>% 
  select(if_arr_delay, arr_delay)
```

Create more variables

```{r}
flights %>% 
  mutate(if_arr_delay = (arr_delay > 0), air_time_hour = air_time/60) %>% 
  select(if_arr_delay, arr_delay, air_time_hour, air_time)
```

**Exercise 3.** Create new variable which tells us what is the sum of departure and arrival delay.


## `filter`

Choose rows that had `dep_delay` greater than 20 mins.

```{r}
flights %>% 
  filter(dep_delay > 20) %>% 
  arrange(dep_delay)
```

Choose rows that had `dep_` **AND** `arr_delay` greater than 20 mins.

```{r}
flights %>% 
  filter(dep_delay > 20, arr_delay > 20) %>% 
  select(dep_delay, arr_delay) %>% 
  arrange(dep_delay, desc(arr_delay))
```

Choose rows that had `dep_` or `arr_delay` greater than 20 mins.

```{r}
flights %>% 
  filter(dep_delay > 20 | arr_delay > 20) %>% 
  select(dep_delay, arr_delay) %>% 
  arrange(dep_delay, desc(arr_delay))
```


Choose rows that had `dep_delay` greater than 20 mins and the plane started from JFK.

```{r}
flights %>% 
  filter(dep_delay > 20, origin == "JFK") %>% 
  select(dep_delay, origin)
```

Choose rows that had `dep_delay` greater than 20 mins and the plane didn't start from JFK.

```{r}
flights %>% 
  filter(dep_delay > 20, origin != "JFK") %>% 
  select(dep_delay, origin)
```

**Exercise 4.** Choose rows that had distance more than 800 miles and arrived in ORD. Then check out the same arrival port but flights that had distance **less** than 800 miles.

## `summarise`

Count mean departure delay

```{r}
flights %>% 
  summarise(dep_delay = mean(dep_delay, na.rm = TRUE))
```


Count mean and median departure delay

```{r}
flights %>% 
  summarise(dep_delay_mean = mean(dep_delay, na.rm = TRUE), 
            dep_delay_median = median(dep_delay, na.rm = TRUE))
```


The same but with the number of rows in the dataset

```{r}
flights %>% 
  filter(!is.na(dep_delay)) %>% 
  summarise(dep_delay_mean = mean(dep_delay, na.rm = TRUE), 
            dep_delay_median = median(dep_delay, na.rm = TRUE),
            n = n())
flights <- flights %>% filter(!is.na(dep_delay))
```


**Exercise 5.** Count mean, median and max arrival delay.


## `group_by`

Count mean departure delay when in groups by origin port

```{r}
flights %>% 
  group_by(origin) %>% 
  summarise(dep_delay_mean = mean(dep_delay))
```


Count how many flights there where arriving in a certain city and departing from a certain city

```{r}
flights %>% 
  group_by(origin, dest) %>% 
  summarise(n = n())
```

Choose which destination had the greatest delay depending on the origin


```{r}
flights %>% 
  group_by(origin) %>% 
  filter(dep_delay == max(dep_delay)) %>% 
  select(origin, dep_delay, dest)
```


**Exercise 6.** Count mean arrival delay by month.

**Exercise 7.** Count max `air_time` by origin.

## `join`

Join information about planes with the flights to know what was the delay by the type of engine

```{r}
data(planes)
flights %>% 
  left_join(planes, by = c("tailnum" = "tailnum")) %>% 
  group_by(year.y) %>% 
  summarise(dep_delay = mean(dep_delay))
```



Join information about planes with the flights to know what was the delay by the number of seats

```{r}
data(planes)
flights %>% 
  left_join(planes, by = c("tailnum" = "tailnum")) %>% 
  group_by(seats) %>% 
  summarise(dep_delay = mean(dep_delay))
```


**Exercise 8.** Join information about planes with the flights to know what was the delay by the manufacturer





