---
title: "R-Ladies dplyr workshop"
output: html_notebook
---

Load libraries `dplyr` and `gapminder`.
```{r}
library(gapminder)
library(dplyr)
```

Our data excerpt of the Gapminder data on life expectancy, GDP per capita, and population by country.

```{r}
gapminder
class(gapminder)
str(gapminder)
```

### Use `filter` to subset data row-wise

```{r}
filter(gapminder, country == "Poland")
```

**pipe** operator `%>%` that `dplyr` imports from `magrittr` package.
The RStudio keyboard shortcut to insert it: `Ctrl + Shift + M` (Windows), `Cmd + Shift + M` (Mac).

```{r}
gapminder %>%
  filter(country == "Poland") -> pl
```

Now lets's modificate the table a bit. Use `mutate` to change the value of `country` column and introduce `NA` there.

```{r}
pl %>%
  mutate(country = NA) -> pl
```

Further let's filter all the Asian countries to a separate table and join `pl` data frome to it.
```{r}
gapminder %>%
  filter(continent == "Asia") -> asia

rbind(pl, asia) -> pl_asia
```


Now let's look at our table `pl_asia` and the behaviour of `filter` with `factors`
```{r}
pl_asia %>% nrow

pl_asia %>% filter(country == "Afghanistan") %>% nrow
pl_asia %>% filter(country != "Afghanistan") %>% nrow
pl_asia %>% filter(!country %in% "Afghanistan") %>% nrow
```

Why the number of rows differs?

Filter using multiple conditions `&`, `|`, `%in%` etc.

```{r}
gapminder %>%
  filter(continent == "Asia", country == "Afghanistan")

gapminder %>%
  filter(continent == "Asia" & country == "Afghanistan")
```

###Use `mutate` to change the value of a column

But we can use it to create a new column.

```{r}
gapminder %>%
  mutate(country_uppercase = toupper(country))
```

This adds a new column `country_uppercase` with country in capital letters.

Now let's set population `pop` values to 0 for `year` 1997 in a table **conditioning** on other column. Use `ifelse` or stricter `if_else`.

```{r}
gapminder %>%
  mutate(population_adjusted = if_else(year == 1997, 0, pop)) -> adj_gapminder
         
adj_gapminder %>%
  filter(year == 1997)
```

### Task for you
Now a task for you. Do the same thing and create **pl_asia** `tibble` without `rbind` and only using `filter` and and `mutate`.

### Use `select` to subset the data on variables or columns.

```{r}
gapminder %>%
  select(country, pop)
```

Select can also `rename` a column

```{r}
gapminder %>%
  select(country, pupulation = pop)
```

Use `select` to displace the variable. There are varius helper functions that works with `select`. See `?everything`.

```{r}
gapminder %>%
  select(pupulation = pop, everything())
```

Now let's report `GDP per capita` relative to `Poland`. Call this new column `gdpPercap_relative_pl`.

* `Filter` down the rows for Poland.
* Create a new temporary `tmp` variable in gapminder:
  * Extract the `gdpPercap` variable from the Polish data.
  * `Replicate` (use `rep`) it once per country in the dataset, so it has the right length.
* Divide raw `gdpPercap` by this Polish figure and create `gdpPercap_relative_pl`.
* Discard the temporary variable of replicated Polish gdpPercap.

```{r}
pl <- gapminder %>%
  filter(country == "Poland")

pl_gap <- gapminder %>%
  mutate(tmp = rep(pl$gdpPercap, nlevels(country)),
         gdpPercap_relative_pl = gdpPercap / tmp,
         tmp = NULL)
```

`mutate` builds new variables sequentially so you can reference earlier ones (like `tmp`) when defining later ones (like `gdpPercapRel`). Also, you can get rid of a variable by setting it to `NULL`.

### Use `arrange` to row-order data in a principled way

Let's arrange data by `year` and by `country`

```{r}
gapminder %>%
  arrange(year, country)
```

Now let's do the same but in descending order and use `desc`

```{r}
gapminder %>%
  arrange(-year, desc(country))
```

### Use `group_by` to report information on a group level.

Let's check how many observations we have on a continent level.
Use `sumamrise` to summarise a single column with count function `n()`.

```{r}
gapminder %>%
  group_by(continent) %>%
  summarise(n = n())
```

Use `count` to count the number of observations that honors grouping.
```{r}
gapminder %>%
  count(continent)
```

Now let's summarise **every** column using `summarise_each` and calculate `min`, `max`, `median`, `mean` for `lifeExp`, `gdpPercap`.

```{r}
gapminder %>%
  group_by(continent) %>%
  summarise_each(funs(min, max, mean, median), lifeExp, gdpPercap)
```

### Task for you

1. For `pl_gap` table filter data to `Europe` and find which country has the largest GDP per capita comparing to Poland (`gdpPercap_relative_pl`) in `year` 2007.

Use `top_n` to select top (or bottom) n rows. 

2. Next which country experienced the sharpest 5-year drop in life expectancy? Recall that this excerpt of the `Gapminder` data only has data every five years, e.g. for 1952, 1957, etc. So this really means looking at life expectancy changes between adjacent timepoints.

### Non-standard evaluation (NSE)
`Dplyr` is using non-standard evaluation, but every function has it's SE version. `group_by` => `group_by_`, `arrage` => `arrange_` etc.

```{r}
columns_to_group <- c("continent")

gapminder %>%
    group_by_(.dots = columns_to_group) %>%
    summarise(n = n())
```

Why this is handy? For example when converting your `dplyr` statements to functions.

```{r}
my_group_n_foo <- function (gap_data, columns_to_group) {
  gap_data %>%
    group_by_(.dots = columns_to_group) %>%
    summarise(n = n())
}

my_group_n_foo(gapminder, "continent")
my_group_n_foo(gapminder, c("continent", "country"))
```





